---
title: "Classify Eye State Analysis for Eye Detection"
author: "Shaheer Ahsan, Mustafa Ssaid, Shahd Abu-Obeid, Anand Sabnis"
date: "`r Sys.Date()`"
format: pdf
toc: true
---

## Introduction

The purpose of this project is to classify the eye state (open or closed) using EEG data collected from 14 electrodes placed on the scalp. The dataset consists of 10,486 observations with 14 numerical predictors and one binary response variable (eyeDetection). The focus is to evaluate the performance of a Support Vector Machine (SVM) model with a Radial Basis Function (RBF) kernel. This model is chosen for its ability to handle non-linear patterns effectively, which are common in EEG data. Additionally, we will also evaluate the performance of a K-Nearest Neighbors (KNN) model. KNN classifies data based on the majority class of nearby points, with the number of neighbors (k) being a key parameter. Features were scaled to ensure fair distance calculations, and cross-validation was used to select the optimal k. The model's performance was evaluated using error rates, confusion matrices, and visualizations to compare its effectiveness with SVM.

## Methodology/Models

The models that will be implemented for this research project include **Support Vector Machine (SVM)** and **K-Nearest Neighbor (KNN)**

1.  Radial SVM uses the RBF kernel, represented as:

> K(x, x') = exp(-y \|\| x - x '\|\|\^2)

where:

-   y controls the influence of individual points.
-   cost balances classification accuracy and margin width.

The dataset was divided into 80% for training and validation, and 20% for testing. Features were scaled to ensure equal importance. The best hyperparameters were identified through 5-fold cross-validation, optimizing both cost and 𝛾. Performance was assessed using error rates, confusion matrices, and visualizations.

2.  K-Nearest Neighbors (KNN) is a non-parametric, instance-based learning algorithm that classifies data points based on the majority class of their nearest neighbors. The key parameter in KNN is the number of neighbors k, which controls the size of the neighborhood used for classification.

The dataset was divided into 80% for training and validation, and 20% for testing. Features were scaled to ensure equal importance, as KNN relies on distance metrics like Euclidean distance, which are sensitive to the scale of features. The optimal value of k was identified through 5-fold cross-validation by minimizing the validation error rate. Performance was assessed using error rates, confusion matrices, and visualizations.

## Data Analysis

### 1. Support Vector Machine (Shaheer Ahsan & Mustafa Ssaid)

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(e1071)
library(GGally)
```

**Load Data**

```{r}
# Load training and test data
train_data <- read.csv("train.csv")
test_data <- read.csv("test.csv")  # No eyeDetection column

# Convert eyeDetection to a factor
train_data$eyeDetection <- as.factor(train_data$eyeDetection)

# Check data structure
str(train_data)

```

**Exploratory Data Analysis**

```{r}
# Pairwise plot of first few features
ggpairs(train_data[, c(1:5, 15)], aes(color = eyeDetection))

# Correlation between first two electrodes
cor(train_data$AF3, train_data$F7)

```

The correlation between the first two features (AF3 and F7) was found to be weak (𝑟= −0.22), indicating low collinearity. Similar patterns were observed across other features, suggesting that all predictors could contribute to the model without significant redundancy.

**Scaling and Data Splitting**

```{r}
# Scale features
set.seed(123)
train_scaled <- scale(train_data[, -15])
test_scaled <- scale(test_data, center = attr(train_scaled, "scaled:center"), 
                     scale = attr(train_scaled, "scaled:scale"))

# Add eyeDetection back to the scaled training data
train_scaled <- data.frame(train_scaled)
train_scaled$eyeDetection <- train_data$eyeDetection

# Split training data into train (80%) and validation (20%)
n <- nrow(train_scaled)
train_idx <- sample(1:n, 0.8 * n)
train_set <- train_scaled[train_idx, ]
val_set <- train_scaled[-train_idx, ]

```

**Model Tuning**

```{r}
# Tune SVM with radial kernel
set.seed(123)
svm_tune <- tune(svm,
                 eyeDetection ~ .,
                 data = train_set,
                 kernel = "radial",
                 ranges = list(cost = c(1, 10), gamma = c(0.1, 1)),
                 tunecontrol = tune.control(cross = 5))

# Best model summary
summary(svm_tune)

```

The Radial SVM model was tuned over the following ranges:

-   cost: {1, 10}
-   y: {0.1, 1}

*Results:*

-   Best parameters: cost = 10, y = 1

-   Cross-validation error: 5.48%

> This low error rate demonstrates the model's ability to generalize across folds.

**Model Evaluation**

```{r}
# Predict on validation set
val_predictions <- predict(svm_tune$best.model, val_set[, -15])

# Confusion matrix
val_conf_matrix <- table(True = val_set$eyeDetection, Predicted = val_predictions)
val_conf_matrix

# Error rate
val_error_rate <- mean(val_predictions != val_set$eyeDetection)
val_error_rate

```

The validation error rate of **4.77%** highlights the model's strong performance. Most misclassifications occur in borderline cases, which is expected with noisy EEG data.

**Test Set Prediction**

```{r}
# Predict on the test set (no labels available for error computation)
test_predictions <- predict(svm_tune$best.model, test_scaled)

# Bar plot of predicted classes
library(ggplot2)
ggplot(data.frame(Predictions = test_predictions), aes(x = Predictions)) +
  geom_bar(fill = "blue") +
  labs(title = "Distribution of Predicted Classes", x = "Predicted Class", y = "Count")


```

The bar plot shows a balanced distribution of predicted classes (0 and 1), indicating that the model is not biased toward either class.

**Visualization**

```{r}
# Visualize decision boundary using PCA
pca_result <- prcomp(train_scaled[, -15])
pca_train <- data.frame(pca_result$x, eyeDetection = train_scaled$eyeDetection)
ggplot(pca_train, aes(x = PC1, y = PC2, color = eyeDetection)) +
  geom_point() +
  labs(title = "PCA Projection of Training Data", x = "Principal Component 1", 
       y = "Principal Component 2")

```

The PCA plot shows clear clustering of the two classes, supporting the Radial SVM's ability to separate non-linear patterns effectively.

**Conclusion**

The Radial SVM model achieved an error rate of 4.77% on the validation set, demonstrating its strong performance in classifying eye states based on EEG data. The low error rate and balanced predictions highlight the model's generalizability.

### 2. K-Nearest Neighbors (Shahd Abu-Obeid & Anand Sabnis)

```{r setup2, include=FALSE}
library(knitr)
library(ggplot2)
library(class)
```

**Load Data**

```{r}
# Load training and test data
train_data <- read.csv("train.csv")
test_data <- read.csv("test.csv")  # No eyeDetection column

# Convert eyeDetection to a factor
train_data$eyeDetection <- as.factor(train_data$eyeDetection)

# Check data structure
str(train_data)
```

**Exploratory Data Analysis**

```{r}
# Pairwise plot of first few features
ggpairs(train_data[, c(1:5, 15)], aes(color = eyeDetection))

# Correlation between first two electrodes
cor(train_data$AF3, train_data$F7)
```

**Scaling and Data Splitting**

```{r}
# Scale features
set.seed(123)
train_scaled <- scale(train_data[, -15])
test_scaled <- scale(test_data, center = attr(train_scaled, "scaled:center"), 
                     scale = attr(train_scaled, "scaled:scale"))

# Add eyeDetection back to the scaled training data
train_scaled <- data.frame(train_scaled)
train_scaled$eyeDetection <- train_data$eyeDetection

# Split training data into train (80%) and validation (20%)
n <- nrow(train_scaled)
train_idx <- sample(1:n, 0.8 * n)
train_set <- train_scaled[train_idx, ]
val_set <- train_scaled[-train_idx, ]
```

**Model Tuning**

```{r}
# Tune KNN using cross-validation
set.seed(123)
k_values <- c(3, 5, 7, 9)
val_error_rates <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  k <- k_values[i]
  knn_predictions <- knn(
    train = train_set[, -15],
    test = val_set[, -15],
    cl = train_set$eyeDetection,
    k = k
  )
  val_error_rates[i] <- mean(knn_predictions != val_set$eyeDetection)
}

# Optimal K
optimal_k <- k_values[which.min(val_error_rates)]

# Plot validation error rates for different k values
library(ggplot2)
ggplot(data.frame(k = k_values, ErrorRate = val_error_rates), aes(x = k, y = ErrorRate)) +
  geom_line() +
  geom_point() +
  labs(title = "Validation Error Rates for KNN", x = "Number of Neighbors (k)", y = "Error Rate")
```

**Model Evaluation**

```{r}
# Predict on validation set using optimal K
set.seed(123)
val_predictions <- knn(
  train = train_set[, -15],
  test = val_set[, -15],
  cl = train_set$eyeDetection,
  k = optimal_k
)

# Confusion matrix
val_conf_matrix <- table(True = val_set$eyeDetection, Predicted = val_predictions)
val_conf_matrix

# Error rate
val_error_rate <- mean(val_predictions != val_set$eyeDetection)
val_error_rate
```

**Test Set Prediction**

```{r}
# Predict on the test set (no labels available for error computation)
set.seed(123)
test_predictions <- knn(
  train = train_set[, -15],
  test = test_scaled,
  cl = train_set$eyeDetection,
  k = optimal_k
)

# Bar plot of predicted classes
ggplot(data.frame(Predictions = test_predictions), aes(x = Predictions)) +
  geom_bar(fill = "green") +
  labs(title = "Distribution of Predicted Classes for KNN", x = "Predicted Class", y = "Count")
```

**Visualization**

```{r}
# Visualize decision boundary using PCA
pca_result <- prcomp(train_scaled[, -15])
pca_train <- data.frame(pca_result$x, eyeDetection = train_scaled$eyeDetection)
ggplot(pca_train, aes(x = PC1, y = PC2, color = eyeDetection)) +
  geom_point() +
  labs(title = "PCA Projection of Training Data (KNN)", x = "Principal Component 1", 
       y = "Principal Component 2")
```

The PCA plot provides a visual representation of the clustering achieved by the KNN model.

**Conclusion**

The KNN model achieved an error rate of 5.24% on the validation set, demonstrating its effectiveness in classifying eye states based on EEG data. The slightly higher error rate compared to SVM highlights its limitations, but its performance remains competitive and reliable.

## Conclusion

The primary objective of this project was to classify eye states (open or closed) using EEG data with Support Vector Machine (SVM) and K-Nearest Neighbors (KNN) models. The results indicate that both models performed effectively, with SVM achieving a lower validation error rate (4.77%) compared to KNN (5.24%). This highlights the superior ability of SVM with a radial kernel to handle non-linear patterns in EEG data.

During the analysis, challenges such as data scaling and hyperparameter tuning were encountered. While cross-validation helped mitigate overfitting, the noisy nature of EEG signals presented difficulties in achieving perfect separation of classes. Parameter selection, especially for SVM (cost and gamma) and KNN (k-value), required iterative optimization to balance accuracy and computational efficiency.

For future improvements, additional data preprocessing techniques, such as signal denoising or feature selection, could further enhance classification accuracy. Incorporating advanced models like neural networks or ensemble methods, alongside robust hyperparameter search strategies, may yield better results. Moreover, using domain-specific features from EEG signals could provide valuable insights for classification tasks.

In conclusion, the **SVM** model proved to be the most effective in this analysis, showcasing its robustness in dealing with complex and high-dimensional datasets like EEG data.

## References

1.  Chang, C.-C., & Lin, C.-J. (2011). LIBSVM: A library for support vector machines. *ACM Transactions on Intelligent Systems and Technology (TIST), 2(3), 1-27.*
2.  Cover, T., & Hart, P. (1967). Nearest neighbor pattern classification. *IEEE Transactions on Information Theory, 13(1), 21-27.*
3.  Bishop, C. M. (2006). Pattern recognition and machine learning. *Springer*.
4.  Guyon, I., & Elisseeff, A. (2003). An introduction to variable and feature selection. *Journal of Machine Learning Research, 3, 1157-1182.*
5.  Vapnik, V. N. (1995). The nature of statistical learning theory. *Springer Science & Business Media.*
